Stage-Splitting a Modal Language

### 2014-07-24 paper/semantics.tex
few more hat-related fixes

%The evaluation judgment operates on stage-1 code, whereas the
%speculation operates on stage-2 code.  Since $\next$ and $\prev$ are the
%crossover points between 1-code and 2-code, they are correspondingly the only
%places where the evaluation and speculation judgments depend on the other.

%The evaluation judgment is very similar to standard call-by-value evaluation.
%The input to evaluation is a stage-1 expression (usually
%$e$), as well as two administrative contexts ($\Gamma$), covered
%later.  Evaluation has two outputs: the {\em future context} (usually $\xi$) and
%the {\em partial value} (usually $v$).  We cover the latter first.  The partial
%value is essentially the result of the first-stage portion of $e$, and must be a
%\bbone-value of the same type as $e$.  As usual, this means that $v$ must be
%composed only of base primitives, tuples, injections, and lambdas (corresponding
%to base types, products, sums, and functions).  Analogously, the value
%corresponding to $\fut$ is a construct called a {\em hatted variable} (written
%$\hat y$), which signifies a reference to some stage-2 computation.  The mapping
%between these hatted variables and stage-two expressions is
%}}}

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item 
Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully evaluated, 
but which may contain hatted variables. In the example above, 
\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
\item Residuals ($\ressym$es) are stage \bbtwo\ terms whose stage \bbone\
subexpressions have all been fully evaluated. In the example above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
\item Values ($\valsym$s) are stage \bbtwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future environment} $\xi$ and a partial value $v$.
The future environment is a mapping from newly-created variables 
(which then may appear---hatted---in $v$)
to residuals.  It is the formal manifestation of the floating substitutions in our example.
For all of the normal features of \lang (\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and effect on values as standard evaluation,
and the final future environment is gotten by merging the future environments of the subterms.

When this judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it searches into the block's stage \bbtwo\ content to find any contained stage
\bbone\ subexpressions and evaluate them in place.  
We call this search process \emph{speculation}. It is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.
Once the contents of the $\next$ block are speculated into a residual ($q$), 
we then return a fresh hatted variable ($\hat y$), 
along with a future environment which maps that variable to our residual ($y \mapsto q$).

@@

A function $f$ which accepts inputs at both stages \bbone{} and \bbtwo{} can be
given a type of the form $A\to\fut(B\to C)$.%
\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
the following higher-order function which makes the adjustment:
\begin{lstlisting} 
let adjust (f : $int * int -> $int) =
  fn (p : int) => 
    next{
      fn (b : int) => 
        prev{f (next {b}, p)}
    }
\end{lstlisting}}
%
Given its stage \bbone\ argument $a:A$, we can evaluate the partially-applied
function:
$\cdot;\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
ensures that the resulting $f_a$ has type $B\to C$ in the residual language.
Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
the ultimate result of the function, $f_a~b \mathop{\tworedsym} c$.

>>>>

%The evaluation judgment operates on stage-1 code, whereas the
%speculation operates on stage-2 code.  Since $\next$ and $\prev$ are the
%crossover points between 1-code and 2-code, they are correspondingly the only
%places where the evaluation and speculation judgments depend on the other. 
%}}}

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item 
Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully evaluated, 
but which may contain stage-\bbtwo\ variables wrapped in $\next$ blocks. 
In the example above, 
\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
\item Residuals ($\ressym$es) are stage \bbtwo\ terms whose stage \bbone\
subexpressions have all been fully evaluated. In the example above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
\item Values ($\valsym$s) are stage \bbtwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future environment} $\xi$ and a partial value $v$.
The future environment is a mapping from newly-created variables 
(which then may appear within $\next$ blocks in $v$)
to residuals.  It is the formal manifestation of the floating substitutions in our example.
For all of the normal features of \lang (\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and effect on values as standard evaluation,
and the final future environment is gotten by merging the future environments of the subterms.

When this judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it searches into the block's stage \bbtwo\ content to find any contained stage
\bbone\ subexpressions and evaluate them in place.  
We call this search process \emph{speculation}. It is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.
Once the contents of the $\next$ block are speculated into a residual ($q$), 
we then return a fresh variable wrapped in a $\next$ block ($\next~\hat y$), 
along with a future environment which maps that variable to our residual ($\hat y \mapsto q$).

@@

A function $f$ which accepts inputs at both stages \bbone{} and \bbtwo{} can be
given a type of the form $A\to\fut(B\to C)$.%
\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
the following higher-order function which makes the adjustment:
\begin{lstlisting} 
let adjust (f : $int * int -> $int) =
  fn (p : int) => 
    next{
      fn (b : int) => 
        prev{f (next {b}, p)}
    }
\end{lstlisting}}
%
Given its stage \bbone\ argument $a:A$, we can evaluate the partially-applied
function:
$\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
ensures that the resulting $f_a$ has type $B\to C$ in the residual language.
Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
the ultimate result of the function, $f_a~b \mathop{\tworedsym} c$.
